HO unification from object language to meta language

### 2024-02-20 paper/main.tex
more work on intro

\begin{abstract}
  Specifying and implementing a logic from scratch requires significant effort.
  Logical Frameworks and Higher Order Logic Programming Languages provide
  dedicated, high-level Meta Languages (ML) to facilitate this task in two
  key ways: 1) variable binding and substitution are simplified when ML binders
  represent object logic ones; 2) proof construction, and even proof search, is
  greatly simplified by leveraging the unification procedure provided by the ML.
  Notable examples of ML are Elf~\cite{elf}, Twelf~\cite{twelf},
  $\lambda$Prolog~\cite{miller_nadathur_2012} and
  Isabelle~\cite{10.1007/978-3-540-71067-7_7}
  which have been utilized to implement various formal systems such as
  First Order Logic~\cite{felty88cade},
  Set Theory~\cite{10.1007/BF00881873} and
  even Higher Order Logic~\cite{books/sp/NipkowPW02}.

  The object logic we are interested in is Dependent Type Theory (DTT),
  for which we aim to implement a unification procedure \U{} using the ML
  Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
  Elpi comes equipped with the equational theory \Ue{}, comprising
  $\eta\beta$ equivalence
  and higher order unification restricted to the pattern
  fragment~\cite{miller92jsc}.
  We want \U{} to feature the same equational theory as \Ue{} but on the
  object logic DTT.
  Unfortunately the natural encoding of DTT in Elpi, which we refer to as \Fo{},
  ``underuses'' \Ue{} by restricting it to first-order unification problems only. 
  To address this issue, we propose a better-behaved encoding, \Ho{},
  demonstrate how to map unification problems in \Fo{}
  to related problems in \Ho{}, and illustrate
  how to map back the unifiers found by \Ue{}, effectively implementing
   \U{} on top of \Ue{} for the natural encoding \Fo{}.

  We apply this technique to the implementation of a type-class~\cite{wadler89}
  solver for Coq~\cite{Coq-refman}.
  Type-class solvers are proof search procedures based on back-chaining designated
  lemmas, providing essential automation to widely used 
  Coq libraries such as Stdpp/Iris~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
  and TLC~\cite{10.1007/978-3-642-14052-5_15}. These two libraries
  constitute our test bed.
\end{abstract}

@@

% \begin{comment}
% \begin{abstract}

%   % Unification is a fundamental process behind goal resolution in logic
%   % programming. Two terms $t_1$ and $t_2$ unify up to $\eta\beta$-reduction if it
%   % exist a substitution $\theta$ such that $\theta t_1 = \theta t_2$. A
%   % substitution is a mapping from unification variables to terms and it is the most
%   % general, if forall substitution $\theta'$, such that $\theta' t_1 = \theta'
%   % t_2$, it exists a substitution $\theta''$ such that $\theta' = \theta \theta''$.
%   % Unification has been widely studied in the 90s [here the refs]. In this paper we
%   % want to give a framework to reason about unification in meta-programming.
%   Meta-programming \cite{hill1994} is a programming technique in which a
%   program, referred to as meta language (\ml), can treat an other program, knwon
%   as the object language (\ol) as its data. This programming approach is used in
%   various works \cite{cite1, cite2...} to construct logical frameworks (\lf)
%   capable of automating proof resolution. In many of the
%   previously mentioned works, the equational theory expressed by the \ol is
%   weaker than the \ml one. Consequently, no effort is required to adjust the
%   unification algorithm of the \ml to represent the unification process of 
%   the \ol: the former algorithm encompassing the second one. In our
%   specific scenario, however, we work with an \ol with a richer equational
%   theory than our \ml. This paper introduces a framework that enables the \ml to
%   comprehend the unification of the \ol in the context of higher-order 
%   unification. Practical applications of these findings will be provided.

% \end{abstract}
% \end{comment}

@@

Specifying and implementing a logic from scratch requires significant effort.
Logical Frameworks and Higher Order Logic Programming Languages provide
dedicated, high-level Meta Languages (ML) to facilitate this task in two
key ways: 1) variable binding and substitution are simplified when ML binders
represent object logic ones; 2) proof construction, and even proof search, is
greatly simplified by leveraging the unification procedure provided by the ML.
Notable examples of ML are Elf~\cite{elf}, Twelf~\cite{twelf},
$\lambda$Prolog~\cite{miller_nadathur_2012} and
Isabelle~\cite{10.1007/978-3-540-71067-7_7}
which have been utilized to implement various formal systems such as
First Order Logic~\cite{felty88cade},
Set Theory~\cite{10.1007/BF00881873} and
even Higher Order Logic~\cite{books/sp/NipkowPW02}.

The object logic we are interested in is Dependent Type Theory (DTT),
and we want to code a type-class~\cite{wadler89}
solver for Coq~\cite{Coq-refman}.
Type-class solvers are proof search procedures based on back-chaining designated
lemmas, providing essential automation to widely used 
Coq libraries such as Stdpp/Iris~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
and TLC~\cite{10.1007/978-3-642-14052-5_15}. These two libraries
constitute our test bed.

For example, these 3 designated lemmas (Instances in the TC slang)
are supposed to be animated, backward chained, by the TC solver.
fin n is the type of numbers smaller than n, nfact the number of prime
factors and Finite/Decision properties designated as classes what are
automatically proved by Coq using the designated instances.

@@

\noindent For example here

@@

A natural encoding (others are discussed in section xx)
uses c for constants, all for forall, and then app and lam
for the lambda calculus part.

\begin{elpicode}
finite (app[c"fin", N]).
decision (app [c"nfact",N,NF]).
decision (all A x\ app[P, x]) :- 
  (pi x\ decision (app[P, x])), finite A.
\end{elpicode}

Unfortunately this does not work since the last rule cannt
backchain on the encoded goal

\begin{elpicode}
decision (all (app[c"fin",c"7"]) y\ app[c"nfact", y, c"3"]).
\end{elpicode}

indeed this unif problem has no solution

\begin{elpicode}
app[c"nfact", y, c"3"] = app[P, y].
\end{elpicode}

In this paper we study a more sophisticated encoding

\begin{elpicode}
decision (all A x\ Pm x) :- link Pm A P, 
  (pi x\ decision (app[P, x])), finite A.
\end{elpicode}

This time Pm is HO for elpi, an the extra link is there to explain how to
get P out of it.

\begin{elpicode}
app[c"nfact", y, c"3"] = Pm y
Pm = y\app[c"nfact", y, c"3"]
P = lam (app[c"fin",c"7"]) a\ app[c"nfact", a, c"3"]
\end{elpicode}

We also see that this is not trivial, because in the rec call
we have now a beta redex.

\subsection{Alternative encodings} %%%%%%%%%%%%%%%%%%%%%%

>>>>

\begin{abstract}
  Specifying and implementing a logic from scratch requires significant effort.
  Logical Frameworks and Higher Order Logic Programming Languages provide
  dedicated, high-level Meta Languages (ML) to facilitate this task in two
  key ways: 1) variable binding and substitution are simplified when ML binders
  represent object logic ones; 2) proof construction, and even proof search, is
  greatly simplified by leveraging the unification procedure provided by the ML.
  Notable examples of ML are Elf~\cite{elf}, Twelf~\cite{twelf},
  $\lambda$Prolog~\cite{miller_nadathur_2012} and
  Isabelle~\cite{10.1007/978-3-540-71067-7_7}
  which have been utilized to implement various formal systems such as
  First Order Logic~\cite{felty88cade},
  Set Theory~\cite{10.1007/BF00881873},
  Higher Order Logic~\cite{books/sp/NipkowPW02}, and even the Calculus of
  Constuctions~\cite{felty93lics}.

  The object logic we are interested in is Coq's~\cite{Coq-refman}
  Dependent Type Theory (DTT),
  for which we aim to implement a unification procedure \U{} using the ML
  Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
  Elpi comes equipped with the equational theory \Ue{}, comprising
  $\eta\beta$ equivalence
  and higher order unification restricted to the pattern
  fragment~\cite{miller92jsc}.
  We want \U{} to feature the same equational theory as \Ue{} but on the
  object logic DTT. Elpi also comes with an encoding for DTT that works well
  for meta-programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
  Unfortunately this encoding, which we refer to as \Fo{},
  ``underuses'' \Ue{} by restricting it to first-order unification problems only. 
  To address this issue, we propose a better-behaved encoding, \Ho{},
  demonstrate how to map unification problems in \Fo{}
  to related problems in \Ho{}, and illustrate
  how to map back the unifiers found by \Ue{}, effectively implementing
   \U{} on top of \Ue{} for the encoding \Fo{}.

  We apply this technique to the implementation of a type-class~\cite{wadler89}
  solver for Coq~\cite{Coq-refman}.
  Type-class solvers are proof search procedures based on
  unification that back-chain designated lemmas, providing essential
  automation to widely used 
  Coq libraries such as Stdpp/Iris~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
  and TLC~\cite{10.1007/978-3-642-14052-5_15}. These two libraries
  constitute our test bed.
\end{abstract}

@@

@@

Specifying and implementing a logic from scratch requires significant effort.
Logical Frameworks and Higher Order Logic Programming Languages provide
dedicated, high-level Meta Languages (ML) to facilitate this task in two
key ways: 1) variable binding and substitution are simplified when ML binders
represent object logic ones; 2) proof construction, and even proof search, is
greatly simplified by leveraging the unification procedure provided by the ML.
Notable examples of ML are Elf~\cite{elf}, Twelf~\cite{twelf},
$\lambda$Prolog~\cite{miller_nadathur_2012} and
Isabelle~\cite{10.1007/978-3-540-71067-7_7}
which have been utilized to implement various formal systems such as
First Order Logic~\cite{felty88cade},
Set Theory~\cite{10.1007/BF00881873},
Higher Order Logic~\cite{books/sp/NipkowPW02}, and even the Calculus of
Constuctions~\cite{felty93lics}.

The object logic we are interested in is Coq's~\cite{Coq-refman}
Dependent Type Theory (DTT), and we want to code a type-class~\cite{wadler89}
solver for Coq~\cite{Coq-refman} using the Coq-Elpi~\cite{tassi:hal-01637063}
meta programming framework.
Type-class solvers are unification based proof search procedures
that combine a set of designated lemmas in order to providing essential
automation to widely used Coq libraries.

As the running example we take the \coqIn{Decide} type class,
from the Stdpp~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
library. The class identifies predicates equipped with a decision procedure.
The following three designated lemmas (called \coqIn{Instances} in the
type-class jargon) state that: 1) the type \coqIn{fin n}, of natural numbers
smaller than \coqIn{n} is finite; 2) the predicate \coqIn{nfact n nf},
linking a natural number \coqIn{n} to its prime factors \coqIn{nf}, is decidable;
3) the universal closure of a predicate has a decision procedure if the
predicate has and if its domain is finite.

@@

\noindent Under this context the type-class solver is able to prove
the the following statement automatically by back-chaining
the three instances.

@@

\noindent
The encoding of DTT provided by Elpi, that we will discuss at length later in
section~\ref{sec:encodings} and ~\ref{} features the following term constructors:

\begin{elpicode}
kind tm type.
type lam tm -> (tm -> tm) -> tm. % lambda abstraction
type app list tm -> tm.          % n-ary application
type all tm -> (tm -> tm) -> tm. % forall quantifier
type c string -> tm.             % constants
\end{elpicode}

\noindent
\marginpar{TODO: explain HOAS}
Following this term encoding the three instances are represented by the
following rules:

\begin{elpicode}
finite (app[c"fin", N]).
decision (app [c"nfact", N, NF]).
decision (all A x\ app[P, x]) :- finite A,
  pi x\ decision (app[P, x]).
\end{elpicode}

\noindent
\marginpar{TODO: explain pi, cons}
Unfortunately this direct translation of the instances considers the
predicate \coqIn{P} as a first order term. If we try to backchain the
third rule on the encoding of the goal above:

\begin{elpicode}
decision (all (app[c"fin", c"7"]) y\
  app[c"nfact", y, c"3"]).
\end{elpicode}

\noindent
we fail because of this ``higher order'' unification problem (in DTT)
is phrased as a first order unification problem in the meta language.

\begin{elpicode}
app[c"nfact", y, c"3"] = app[P, y]
\end{elpicode}

\noindent
In this paper we study a more sophisticated encoding of Coq terms allowing
us to rephrase the problematic rule as follows:

\begin{elpicode}
decision (all A x\ Pm x) :- link Pm A P, finite A,
  pi x\ decision (app[P, x]).
\end{elpicode}

\noindent
This time \elpiIn{Pm} is an higher order unification variable (of type
\elpiIn{tm -> tm}). The resulting unification problem is now:

\begin{elpicode}
app[c"nfact", y, c"3"] = Pm y
\end{elpicode}

\noindent
That admits one solution:

\begin{elpicode}
Pm = y\ app[c"nfact", y, c"3"]
A = app[c"fin",c"7"]
\end{elpicode}

\noindent
Elpi succeeds in the application of the new rule and then runs
the premise \elpiIn{link Pm A P} that is in charge of bringing the
assignment back to the domain of Coq terms (the type \elpiIn{tm}):

\begin{elpicode}
P = lam A a\ app[c"nfact", a, c"3"]
\end{elpicode}

\noindent
This simple example is sufficient to show that the encoding we seek
is not trivial. Indeed the solution for \elpiIn{P} generates a
(Coq) $\beta$-redex in the second premise (under the \elpiIn{pi x}):

\begin{elpicode}
decision (app[lam A (a\ app[c"nfact", a, c"3"]), x])
\end{elpicode}

\noindent
In turn the redex prevents the second rule to backchain properly since
the following unification problem has no solution:

\begin{elpicode}
app[lam A (a\ app[c"nfact", a, c"3"]), x] =
app[c"nfact", N, NF]
\end{elpicode}

\noindent
This time the root cause is that the unification procedure of \Ue{} of the
meta language is not aware of the equational theory of the object logic \U{},
even if both theories include $\eta\beta$-conversion and admit most general
unifiers for problems in the pattern fragment~\cite{miller92jsc}.

In this paper we discuss alternative encodings of Coq in
Elpi~\ref{sec:encodings}, then we identify a minimal language \Fo{}
in which the problems sketched here can be fully described.
We then detail an encoding \elpiIn{comp} from \Fo{} to \Ho{} (the language of
the meta language) and a decoding \elpiIn{decomp} to relate the unifiers
bla bla..

\section{Alternative encodings} %%%%%%%%%%%%%%%%%%%%%%
\label{sec:encodings}
